---
title: "IPL 2026 Winner Prediction Utilising Machine Learning and Monte Carlo Simulation"
author: "Tej Mehta"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(janitor)
library(glue)
library(tidymodels)   # unified modelling framework
library(xgboost)      # gradient boosting
library(ranger)       # random forest package
library(glmnet)       # logistic regression
library(probably)     # probability calibration
library(ggplot2)
library(scales)
library(ggtext)
library(patchwork)
library(gt)
library(gtExtras)
library(here)
library(tictoc)
library(furrr)      # nice package to use to speed up rolling calculations
library(vip)

##################################### Setup ######################################

# Project constants
current_season <- "2026"
historical_data <- read.csv("~/Downloads/IPL.csv")
matches_2026 <- read.csv("~/Downloads/archive (7)/matches.csv")
deliveries_2026 <- read.csv("~/Downloads/archive (7)/deliveries.csv")
squads_2026 <- read.csv("~/Downloads/archive (7)/squads.csv")
venues_2026 <- read.csv("~/Downloads/archive (7)/venues.csv")
points_2026 <- read.csv("~/Downloads/archive (7)/points_table.csv")

## Team names standardisation
team_name_map <- c(
  "Delhi Daredevils" = "Delhi Capitals",
  "Kings XI Punjab" = "Punjab Kings",
  "Royal Challengers Bangalore" = "Royal Challengers Bengaluru",
  "Rising Pune Supergiant" = "Rising Pune Supergiants",
  "Deccan Chargers" = "Deccan Chargers",                # defunct (old team)
  "Kochi Tuskers Kerala" = "Kochi Tuskers Kerala",      # defunct (old team)
  "Pune Warriors" = "Pune Warriors",                    # defunct (old team)
  "Gujarat Lions" = "Gujarat Lions"                     # defunct (old team)
)

# Active 2026 teams
active_teams_2026 <- c("Chennai Super Kings", "Delhi Capitals", "Gujarat Titans",
  "Kolkata Knight Riders", "Lucknow Super Giants", "Mumbai Indians", "Punjab Kings",
  "Rajasthan Royals", "Royal Challengers Bengaluru", "Sunrisers Hyderabad")

# Colour palette for data visualisation
team_colours <- c(
  "Chennai Super Kings" = "#FDB913",
  "Delhi Capitals" = "#0078BC",
  "Gujarat Titans" = "#0C192D",
  "Kolkata Knight Riders" = "#3A225D",
  "Lucknow Super Giants" = "#7C1C1C",
  "Mumbai Indians" = "#003FA0",
  "Punjab Kings" = "#C9A227",
  "Rajasthan Royals" = "#D415DA",
  "Royal Challengers Bengaluru" = "#EC1C24",
  "Sunrisers Hyderabad" = "#FF822A"
)

##################################### Data prep ######################################
# Ensuring historical data columns are correct dtype
historical_data <- read_csv("~/Downloads/IPL.csv",
  col_types = cols(
    match_id = col_character(),
    date = col_date(format = "%Y-%m-%d"),
    season = col_character(),
    innings = col_integer(),
    over = col_integer(),
    ball = col_integer(),
    runs_batter = col_integer(),
    runs_extras = col_integer(),
    runs_total = col_integer(),
    valid_ball = col_integer()
  )
)

### Harmonise team names
harmonise_teams <- function(x) {
  recode(x, !!!team_name_map)
}

deliveries_clean <- historical_data %>%
  mutate(
    batting_team = harmonise_teams(batting_team),
    bowling_team = harmonise_teams(bowling_team),
    toss_winner = harmonise_teams(toss_winner),
    match_won_by = harmonise_teams(match_won_by),
    season_year = {
      first_year <- as.integer(str_extract(season, "^\\d{4}"))
      is_split <- str_detect(season, "/")
      if_else(is_split, first_year + 1L, first_year)
    }
  )

# Match level table
## This match level table will help extract distinct match records for efficient modelling later.
## The historical ball-by-ball file contains match metadata repeated on every row.

matches_clean <- deliveries_clean %>%
  group_by(match_id) %>%
  slice(1) %>%              # 1 row per match
  ungroup() %>%
  select(
    match_id, date, season, season_year, venue, city, batting_team, bowling_team,
    toss_winner, toss_decision, match_won_by, win_outcome, stage, event_match_no, 
    player_of_match, balls_per_over, overs) %>%
  rename(
    team1 = batting_team,
    team2 = bowling_team) %>%
  mutate(
    bat_first = team1, # which team batted first
    bat_second = team2,
    toss_bat = toss_decision == "bat", # did toss winner choose to bat first
    winner = match_won_by,
    team1_won = as.integer(winner == team1), # binary win/no win for team 1 (useful for some models)
    has_result = !is.na(win_outcome) & win_outcome != "no result"
  )

# DQ checks
## Fixed the 2007/2008, 2009/2010, 2020/2021 seasons which are not getting picked up by the season and is coming up as NA
message("\n--- Season coverage ---")
matches_clean %>%
  count(season_year, season) %>%
  print(n = 30)

message("\n--- Null check on key modelling columns ---")
matches_clean %>%
  select(match_id, date, team1, team2, toss_winner, toss_decision, winner) %>%
  summarise(across(everything(), ~sum(is.na(.)))) |>
  print()

message("\n--- Active teams appearing in 2025/2026 ---")
matches_clean %>%
  filter(season_year >= 2025) %>%
  distinct(team1) %>%
  arrange(team1) %>%
  print()

##################################### EDA ######################################
### Helper: consistent ggplot theme for this project
theme_ipl <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_markdown(face = "bold", size = 14),
      plot.subtitle = element_text(colour = "grey40", size = 11),
      plot.caption = element_text(colour = "grey60", size = 9),
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

# Season win counts
season_winners <- matches_clean %>%
  filter(has_result, !is.na(winner)) %>%
  count(winner, name = "wins") %>%
  arrange(desc(wins)) %>%
  mutate(
    winner = fct_reorder(winner, wins),
    is_active = winner %in% active_teams_2026,
    bar_colour = if_else(is_active, team_colours[as.character(winner)], "grey70")
  )

p1 <- season_winners |>
  ggplot(aes(x = winner, y = wins, fill = I(bar_colour))) +
  geom_col(width = 0.7) +
  geom_text(aes(label = wins), hjust = -0.2, size = 3.5) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "**IPL all-time match wins** (2008–present)",
    subtitle = "Greyed bars = defunct teams",
    x = NULL,
    y = "Matches won",
  ) +
  theme_ipl()

# Toss advantage (likelihood of winning the game when toss is won)
toss_analysis <- matches_clean %>%
  filter(has_result, !is.na(winner)) %>%
  mutate(toss_winner_won = toss_winner == winner) %>%
  group_by(season_year) %>%
  summarise(
    matches = n(),
    toss_winner_won = sum(toss_winner_won),
    pct = toss_winner_won / matches
  )

p2 <- toss_analysis |>
  ggplot(aes(x = season_year, y = pct)) +
  geom_line(colour = "#004BA0", linewidth = 1) +
  geom_point(colour = "#004BA0", size = 2.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(labels = percent_format(), limits = c(0.3, 0.7)) +
  scale_x_continuous(breaks = seq(2008, 2026, 2)) +
  labs(
    title = "**Toss win → match win rate** by season",
    subtitle = "Dashed line = 50% (no toss advantage)",
    x = "Season",
    y = "% of toss winners who won the match"
  ) +
  theme_ipl()

# Toss decision split (bat first vs field first over time)
toss_decision <- matches_clean %>%
  filter(!is.na(toss_decision)) %>%
  count(season_year, toss_decision) %>%
  group_by(season_year) %>%
  mutate(pct = n / sum(n))

p3 <- toss_decision %>%
  ggplot(aes(x = season_year, y = pct, fill = toss_decision)) +
  geom_area(alpha = 0.85, position = "stack") +
  scale_fill_manual(values = c("bat" = "#F7A721", "field" = "#0078BC"),
                    labels = c("bat" = "Chose to bat", "field" = "Chose to field")) +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(breaks = seq(2008, 2026, 2)) +
  labs(
    title = "**Toss decision trends** — bat first vs field first",
    x = "Season",
    y = "Proportion of toss winners",
    fill = NULL
  ) +
  theme_ipl()

# Average 1st innings score by season
innings1_scores <- deliveries_clean %>%
  filter(innings == 1, valid_ball == 1) %>%
  group_by(match_id, season_year) %>%
  summarise(total_runs = sum(runs_total), .groups = "drop")

p4 <- innings1_scores %>%
  group_by(season_year) %>%
  summarise(
    avg_score = mean(total_runs),
    median_score = median(total_runs),
    q25 = quantile(total_runs, 0.25),
    q75 = quantile(total_runs, 0.75)
  ) %>%
  ggplot(aes(x = season_year)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#EC1C24", alpha = 0.15) +
  geom_line(aes(y = avg_score), colour = "#EC1C24", linewidth = 1) +
  geom_point(aes(y = avg_score), colour = "#EC1C24", size = 2.5) +
  scale_x_continuous(breaks = seq(2008, 2026, 2)) +
  labs(
    title    = "**Average 1st innings total** by season",
    subtitle = "Shaded band = interquartile range",
    x        = "Season",
    y        = "Runs scored (1st innings)"
  ) +
  theme_ipl()

# Head-to-head matrix for active 2026 teams
h2h <- matches_clean %>%
  filter(
    has_result,
    team1 %in% active_teams_2026,
    team2 %in% active_teams_2026,
    !is.na(winner)
  ) %>%
  mutate(
    # Normalise so every pairing appears once per direction
    win_team1 = as.integer(winner == team1)
  ) %>%
  group_by(team1, team2) %>%
  summarise(
    matches  = n(),
    team1_wins = sum(win_team1),
    win_pct  = team1_wins / matches,
    .groups  = "drop"
  )

# Shorten team names for the matrix axis
short_names <- c(
  "Chennai Super Kings" = "CSK",
  "Delhi Capitals" = "DC",
  "Gujarat Titans" = "GT",
  "Kolkata Knight Riders" = "KKR",
  "Lucknow Super Giants" = "LSG",
  "Mumbai Indians" = "MI",
  "Punjab Kings" = "PBKS",
  "Rajasthan Royals" = "RR",
  "Royal Challengers Bengaluru" = "RCB",
  "Sunrisers Hyderabad" = "SRH"
)

# NOTE THE ABOVE H2H IS INCORRECT #

# Build base pairwise H2H table
h2h_base <- matches_clean %>%
  filter(
    has_result,
    !is.na(winner),
    team1 %in% active_teams_2026,
    team2 %in% active_teams_2026
  ) %>%
  mutate(
    team_a = pmin(team1, team2),
    team_b = pmax(team1, team2),
    team_a_won = winner == team_a
  ) %>%
  group_by(team_a, team_b) %>%
  summarise(
    matches = n(),
    team_a_wins = sum(team_a_won),
    team_a_win_pct = team_a_wins / matches,
    .groups = "drop"
  )

h2h_correct <- bind_rows(
  h2h_base %>%
    transmute(
      team1 = team_a,
      team2 = team_b,
      matches = matches,
      wins = team_a_wins,
      win_pct = team_a_win_pct
    ),
  
  h2h_base %>%
    transmute(
      team1 = team_b,
      team2 = team_a,
      matches = matches,
      wins = matches - team_a_wins,
      win_pct = 1 - team_a_win_pct
    )
)

# Heatmap (Correct version)
p5 <- h2h_correct %>%
  mutate(
    t1 = recode(team1, !!!short_names),
    t2 = recode(team2, !!!short_names)
  ) %>%
  ggplot(aes(x = t2, y = t1, fill = win_pct)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(
    aes(label = sprintf("%.0f%%\n(%d)", win_pct * 100, matches)),
    size = 2.8,
    colour = "white",
    fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "#004BA0",
    mid = "grey85",
    high = "#EC1C24",
    midpoint = 0.5,
    labels = scales::percent_format()
  ) +
  labs(
    title = "**Head-to-head win % matrix** (active teams, all time)",
    subtitle = "Each cell shows row team's win % against column team",
    x = "Opponent (column team)",
    y = "Team (row)",
    fill = "Win %"
  ) +
  theme_ipl() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Home advantage analysis
# The IPL.csv venue strings are inconsistent (e.g. "M Chinnaswamy Stadium" vs
# "M. Chinnaswamy Stadium", sometimes with city appended, sometimes not).
# The most robust approach: extract the city from the venue string and join
# to venues_2026 on city, then resolve the home team from there.

venues_map <- venues_2026|>
  clean_names() |>
  # Expand: each city maps to its home team (abbreviation → full name)
  select(city, home_team) |>
  # home_team in venues_2026 is stored as abbreviation (e.g. "MI") — map to full name
  mutate(home_team_full = recode(home_team,
                                 "MI" = "Mumbai Indians",
                                 "CSK" = "Chennai Super Kings",
                                 "KKR" = "Kolkata Knight Riders",
                                 "RCB" = "Royal Challengers Bengaluru",
                                 "DC" = "Delhi Capitals",
                                 "PBKS" = "Punjab Kings",
                                 "RR" = "Rajasthan Royals",
                                 "SRH" = "Sunrisers Hyderabad",
                                 "GT" = "Gujarat Titans",
                                 "LSG" = "Lucknow Super Giants"
  )) |>
  select(city, home_team = home_team_full) |>
  # Some cities appear under multiple spellings in IPL.csv — add aliases
  bind_rows(tibble(
    city = c("Bengaluru", "Mohali", "Visakhapatnam", "Navi Mumbai",
                  "Pune", "Raipur", "Ranchi", "Cuttack"),
    home_team = c("Royal Challengers Bengaluru", "Punjab Kings", "Sunrisers Hyderabad",
                  "Mumbai Indians", "Mumbai Indians", "Chennai Super Kings",
                  "Kolkata Knight Riders", "Kolkata Knight Riders")
  ))

# Extract city from venue string: IPL.csv stores venue as
# "Stadium Name, City" or just "Stadium Name"
# The city is the last comma-separated token; if absent, use the `city` column
venue_wins <- matches_clean |>
  filter(has_result, !is.na(winner), season_year >= 2015) |>
  mutate(
    # Pull city from venue string (last token after final comma), else fall back to city col
    venue_city = coalesce(
      str_trim(str_extract(venue, "[^,]+$")),   # last segment after comma
      city
    )
  ) |>
  # Join on city name
  left_join(venues_map, by = c("venue_city" = "city")) |>
  filter(!is.na(home_team), home_team %in% c(team1, team2)) |>
  mutate(home_won = winner == home_team) |>
  group_by(home_team) |>
  summarise(
    home_matches = n(),
    home_wins = sum(home_won),
    home_win_pct = home_wins / home_matches
  ) |>
  filter(home_team %in% active_teams_2026) |>
  mutate(home_team = fct_reorder(home_team, home_win_pct))

p6 <- venue_wins |>
  ggplot(aes(x = home_team, y = home_win_pct,
             fill = I(team_colours[as.character(home_team)]))) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey40") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.9)) +
  geom_text(aes(label = sprintf("%.0f%%", home_win_pct * 100)),
            hjust = -0.3, size = 5.5, fontface = "bold") +
  coord_flip() +
  labs(
    title = "**Home ground win %** (since 2015)",
    subtitle = "Dashed line = 50% baseline",
    x = NULL,
    y = "Home win %"
  ) +
  theme_ipl()

############################ Feature Engineering ###############################
# The purpose of this is to develop predictive features for the match-winner model.

# Features built include:
#   - Rolling team win rate (last 5 and last 10 matches)
#   - Exponentially weighted form (actuarial decay weighting)
#   - Head-to-head win rate (historical)
#   - Toss interaction (won toss AND field first)
#   - Venue win rate per team
#   - ELO ratings (updated after every match)
#   - Team strength index (average runs scored vs conceded, last 10)
#   - Player ratings (using performance so far in 2026 season & auction price per player)

matches_result <- matches_clean %>%
  filter(has_result, !is.na(winner)) %>%
  mutate(venue = str_trim(str_remove(venue, ",.*$"))) %>%
  # Standardise venue names (i.e. we have "Wankhede Stadium, Mumbai")
  # and "Wankhede Stadium", which are treated as 2 separate venues)
  arrange(date, match_id)

## Helper functions
### Rolling win rate - for a vector of binary win indicators (win = 1)
rolling_win_rate <- function(x, n = 5) {
  slider::slide_dbl(x, mean, .before = n, .after = 0, .complete = FALSE) %>%
    lag(1)      # use past data only (exclude the current match)
}

## Exponential decay weighting (actuarial credibility-style)
### More recent matches weighted more heavily. Lambda controls decay speed.
exp_weighted_avg <- function(x, lambda = 0.85) {
  n <- length(x)
  out <- numeric(n)
  for (i in seq_along(x)) {
    if (i == 1) {
      out[i] <- NA_real_
      next
    }
    past_vals <- x[seq_len(i-1)]
    k <- length(past_vals)
    weights <- lambda^(rev(seq_len(k)) - 1)
    out[i] <- sum(weights * past_vals) / sum(weights)
  }
  out
}

# 1. Build a LONG per-team match history
# Each match appears TWICE — once for each team — so we can compute
# team-specific rolling statistics efficiently.

team_history <- bind_rows(
  matches_result |>
    transmute(
      match_id, date, season_year,
      team      = team1,
      opponent  = team2,
      won       = as.integer(winner == team1),
      toss_won  = as.integer(toss_winner == team1),
      venue
    ),
  matches_result |>
    transmute(
      match_id, date, season_year,
      team      = team2,
      opponent  = team1,
      won       = as.integer(winner == team2),
      toss_won  = as.integer(toss_winner == team2),
      venue
    )
) |>
  arrange(date, match_id, team)

# 2. Rolling form features per team
team_features <- team_history |>
  group_by(team) |>
  arrange(date, match_id) |>
  mutate(
    # Simple rolling win rates
    win_rate_last5  = rolling_win_rate(won, n = 5),
    win_rate_last10 = rolling_win_rate(won, n = 10),
    # Exponentially weighted win rate (more actuarial — downweights old games)
    win_rate_exp    = exp_weighted_avg(won, lambda = 0.85),
    # Cumulative win rate in this season
    season_wins_so_far   = cumsum(won) - won,       # exclude current
    season_played_so_far = row_number() - 1,
    season_win_rate_ytd  = if_else(
      season_played_so_far > 0,
      season_wins_so_far / season_played_so_far,
      NA_real_
    )
  ) |>
  ungroup() |>
  select(match_id, team, win_rate_last5, win_rate_last10, win_rate_exp,
         season_win_rate_ytd)

# 3. Head-to-head win rate

h2h_rates <- team_history |>
  arrange(date, match_id) |>
  group_by(team, opponent) |>
  mutate(
    h2h_wins_so_far   = cumsum(won) - won,
    h2h_played_so_far = row_number() - 1,
    h2h_win_rate      = if_else(
      h2h_played_so_far > 0,
      h2h_wins_so_far / h2h_played_so_far,
      0.5        # shrink toward 0.5 (no prior info → assume equal)
    )
  ) |>
  ungroup() |>
  select(match_id, team, opponent, h2h_win_rate)


# 4. Venue win rate per team
venue_rates <- team_history |>
  arrange(date, match_id) |>
  group_by(team, venue) |>
  mutate(
    venue_wins_so_far   = cumsum(won) - won,
    venue_played_so_far = row_number() - 1,
    venue_win_rate      = if_else(
      venue_played_so_far >= 3,                # need at least 3 games
      venue_wins_so_far / venue_played_so_far,
      NA_real_                                 # flag: insufficient data
    )
  ) |>
  ungroup() |>
  select(match_id, team, venue_win_rate)

# 5. ELO rating system
# Classic Elo: K-factor of 32, starting rating 1500
# Updated after every match; we capture the pre-match rating as the feature
elo_update <- function(rating_a, rating_b, won_a, K = 32) {
  expected_a <- 1 / (1 + 10 ^ ((rating_b - rating_a) / 400))
  new_a      <- rating_a + K * (won_a - expected_a)
  new_b      <- rating_b + K * ((1 - won_a) - (1 - expected_a))
  list(new_a = new_a, new_b = new_b)
}

# Initialise all teams at 1500
all_teams  <- unique(c(matches_result$team1, matches_result$team2))
elo_ratings <- setNames(rep(1500, length(all_teams)), all_teams)

elo_records <- vector("list", nrow(matches_result))

for (i in seq_len(nrow(matches_result))) {
  row   <- matches_result[i, ]
  t1    <- row$team1
  t2    <- row$team2
  won_t1 <- as.integer(row$winner == t1)
  
  # Record PRE-MATCH ratings as features
  elo_records[[i]] <- tibble(
    match_id   = row$match_id,
    elo_team1  = elo_ratings[t1],
    elo_team2  = elo_ratings[t2],
    elo_diff   = elo_ratings[t1] - elo_ratings[t2]
  )
  
  # Update ratings
  new_elos          <- elo_update(elo_ratings[t1], elo_ratings[t2], won_t1)
  elo_ratings[t1]   <- new_elos$new_a
  elo_ratings[t2]   <- new_elos$new_b
}

elo_df <- bind_rows(elo_records)
message("ELO ratings computed for all matches.")

# 6. Innings scoring strength (avg runs scored & conceded, last 10 matches)
innings_strength <- deliveries_clean |>
  filter(innings %in% 1:2, valid_ball == 1) |>
  group_by(match_id, batting_team, bowling_team, innings) |>
  summarise(runs = sum(runs_total), .groups = "drop") |>
  # Join to get dates for ordering
  left_join(matches_result |> select(match_id, date), by = "match_id") |>
  arrange(date, match_id)

batting_strength <- innings_strength |>
  group_by(batting_team) |>
  arrange(date) |>
  mutate(
    avg_runs_scored_last10 = slider::slide_dbl(runs, mean, .before = 10, .after = 0) |> lag(1)
  ) |>
  ungroup() |>
  select(match_id, batting_team, avg_runs_scored_last10)

bowling_strength <- innings_strength |>
  group_by(bowling_team) |>
  arrange(date) |>
  mutate(
    avg_runs_conceded_last10 = slider::slide_dbl(runs, mean, .before = 10, .after = 0) |> lag(1)
  ) |>
  ungroup() |>
  select(match_id, bowling_team, avg_runs_conceded_last10)

# 7. Assemble model dataset — join all features onto match table
# Features for team1
team1_features <- team_features |>
  rename_with(~paste0(.x, "_t1"), -c(match_id, team)) |>
  rename(team1 = team)

team2_features <- team_features |>
  rename_with(~paste0(.x, "_t2"), -c(match_id, team)) |>
  rename(team2 = team)

# Rename opponent so we can join on both team columns — this ensures we pull
# the H2H rate for the specific team1 vs team2 matchup in each match
h2h_t1 <- h2h_rates |>
  rename(team1 = team, team2 = opponent, h2h_win_rate_t1 = h2h_win_rate)

h2h_t2 <- h2h_rates |>
  rename(team2 = team, team1 = opponent, h2h_win_rate_t2 = h2h_win_rate)

venue_t1 <- venue_rates |>
  rename(team1 = team, venue_win_rate_t1 = venue_win_rate)

venue_t2 <- venue_rates |>
  rename(team2 = team, venue_win_rate_t2 = venue_win_rate)

batting_t1 <- batting_strength |>
  rename(team1 = batting_team, avg_runs_scored_t1 = avg_runs_scored_last10)

batting_t2 <- batting_strength |>
  rename(team2 = batting_team, avg_runs_scored_t2 = avg_runs_scored_last10)

bowling_t1 <- bowling_strength |>
  rename(team1 = bowling_team, avg_runs_conceded_t1 = avg_runs_conceded_last10)

bowling_t2 <- bowling_strength |>
  rename(team2 = bowling_team, avg_runs_conceded_t2 = avg_runs_conceded_last10)

model_data <- matches_result |>
  left_join(team1_features, by = c("match_id", "team1")) |>
  left_join(team2_features, by = c("match_id", "team2")) |>
  left_join(h2h_t1 |> select(match_id, team1, team2, h2h_win_rate_t1),
            by = c("match_id", "team1", "team2")) |>
  left_join(venue_t1 |> select(match_id, team1, venue_win_rate_t1),
            by = c("match_id", "team1")) |>
  left_join(venue_t2 |> select(match_id, team2, venue_win_rate_t2),
            by = c("match_id", "team2")) |>
  left_join(batting_t1, by = c("match_id", "team1")) |>
  left_join(batting_t2, by = c("match_id", "team2")) |>
  left_join(bowling_t1, by = c("match_id", "team1")) |>
  left_join(bowling_t2, by = c("match_id", "team2")) |>
  left_join(elo_df, by = "match_id") |>
  mutate(
    # Toss interaction feature: won toss AND chose to field (historically advantageous)
    toss_field_advantage = as.integer(toss_winner == team1 & toss_decision == "field"),
    # Difference features (actuarial: relative risk is more stable than absolute)
    elo_diff             = elo_team1 - elo_team2,
    win_rate_diff_last5  = win_rate_last5_t1 - win_rate_last5_t2,
    win_rate_diff_exp    = win_rate_exp_t1 - win_rate_exp_t2,
    h2h_advantage        = h2h_win_rate_t1 - 0.5,   # centred: positive = team1 historically beats team2
    net_scoring_diff     = (avg_runs_scored_t1 - avg_runs_conceded_t2) -
      (avg_runs_scored_t2 - avg_runs_conceded_t1),
    # Outcome variable for modelling
    outcome              = factor(team1_won, levels = c(0, 1), labels = c("team2_wins", "team1_wins"))
  )

# Auction price
IPL_auction_prices <- read.csv("~/Desktop/IPL Predictive Model/Datasets/IPL_auction_prices.csv")

# Batting rating
batting_stats <- deliveries_clean %>%
  group_by(batter) %>%
  summarise(
    runs = sum(runs_batter, na.rm = TRUE),
    balls = sum(valid_ball, na.rm = TRUE),
    dismissals = sum(striker_out == 1, na.rm = TRUE),
    avg = runs / pmax(dismissals, 1),
    strike_rate = 100 * runs / pmax(balls, 1),
    .groups = "drop"
  ) %>%
  mutate(
    # Credibility: low-ball players get shrunk heavily
    batting_credibility = pmin(balls / 300, 1),
    
    raw_batting_rating =
      0.45 * as.numeric(scale(avg)) +
      0.35 * as.numeric(scale(strike_rate)) +
      0.20 * as.numeric(scale(log1p(runs))),
    
    batting_rating = batting_credibility * raw_batting_rating
  )

# Bowling rating
bowling_stats <- deliveries_clean %>%
  group_by(bowler) %>%
  summarise(
    runs_conceded = sum(runs_bowler, na.rm = TRUE),
    balls = sum(valid_ball, na.rm = TRUE),
    overs = balls / 6,
    wickets = sum(bowler_wicket, na.rm = TRUE),
    economy = runs_conceded / pmax(overs, 1),
    bowling_avg = runs_conceded / pmax(wickets, 1),
    bowling_sr = balls / pmax(wickets, 1),
    .groups = "drop"
  ) %>%
  mutate(
    # Credibility: low-over bowlers get shrunk heavily
    bowling_credibility = pmin(overs / 50, 1),
    
    raw_bowling_rating =
      0.35 * -as.numeric(scale(economy)) +
      0.35 * -as.numeric(scale(bowling_avg)) +
      0.20 * -as.numeric(scale(bowling_sr)) +
      0.10 *  as.numeric(scale(log1p(wickets))),
    
    bowling_rating = bowling_credibility * raw_bowling_rating
  )

# Combine batting + bowling ratings into one performance rating
player_ratings <- full_join(
  batting_stats %>%
    select(player = batter, runs, balls, avg, strike_rate, batting_rating),
  bowling_stats %>%
    select(player = bowler, overs, wickets, economy, bowling_avg, bowling_sr, bowling_rating),
  by = "player"
) %>%
  mutate(
    across(c(batting_rating, bowling_rating), ~ replace_na(.x, 0)),
    
    performance_rating = batting_rating + bowling_rating
  )

# Clean auction data and combine with performance rating
auction_data <- IPL_auction_prices %>%
  rename(
    player = Players,
    sold_price = Sold
  ) %>%
  mutate(
    sold_price = as.numeric(sold_price),
    auction_rating = as.numeric(scale(log1p(sold_price)))
  )

# Final player rating = performance + auction prior
final_player_ratings <- player_ratings %>%
  left_join(
    auction_data %>% select(player, sold_price, auction_rating),
    by = "player"
  ) %>%
  mutate(
    sold_price = replace_na(sold_price, 0),
    auction_rating = replace_na(auction_rating, 0),
    final_rating = 0.7 * performance_rating + 0.3 * auction_rating
  ) %>%
  arrange(desc(final_rating))

# Team strength
squads_2026 <- squads_2026 %>%
  mutate(
    team_name = recode(team_name,
                       "Gujrat Titans" = "Gujarat Titans")
  )

team_strength <- squads_2026 %>%
  left_join(final_player_ratings, by = "player") %>%
  group_by(team_name) %>%
  summarise(
    squad_strength = sum(final_rating, na.rm = TRUE),
    avg_strength = mean(final_rating, na.rm = TRUE),
    n_players = n(),
    .groups = "drop"
  )

################################## Modelling ######################################
# The 3 models that are going to be fit and compared are:
# 1. logistic regression (GLM baseline)
# 2. Random forest (captures non-linear relationships)
# 3. XGBoost (gradient boosting)

# Validation strategy: walk-forward time split (respect time order)

# Build 2026 squad/team strength from final_player_ratings
squad_strength_2026 <- squads_2026 %>%
  clean_names() %>%
  rename(
    team = team_name,
    player = player
  ) %>%
  mutate(
    team = recode(team,
                  "Gujrat Titans" = "Gujarat Titans",
                  "Royal Challengers Bangalore" = "Royal Challengers Bengaluru",
                  "Kings XI Punjab" = "Punjab Kings"
    )
  ) %>%
  left_join(final_player_ratings, by = "player") %>%
  group_by(team) %>%
  arrange(desc(final_rating)) %>%
  slice_head(n = 15) %>%
  summarise(
    squad_strength = sum(final_rating, na.rm = TRUE),
    avg_player_strength = mean(final_rating, na.rm = TRUE),
    n_players_used = n(),
    .groups = "drop"
  )

# Add team strength into model_data
team_strength_t1 <- squad_strength_2026 %>%
  rename(
    team1 = team,
    strength_t1 = squad_strength,
    avg_strength_t1 = avg_player_strength
  )

team_strength_t2 <- squad_strength_2026 %>%
  rename(
    team2 = team,
    strength_t2 = squad_strength,
    avg_strength_t2 = avg_player_strength
  )

model_data <- model_data %>%
  # Remove ANY old strength columns first
  select(-matches("strength"), -matches("n_players_used")) %>%
  
  # Join team1 strength
  left_join(
    team_strength_t1 %>%
      select(team1, strength_t1 = strength_t1, avg_strength_t1 = avg_strength_t1),
    by = "team1"
  ) %>%
  
  # Join team2 strength
  left_join(
    team_strength_t2 %>%
      select(team2, strength_t2 = strength_t2, avg_strength_t2 = avg_strength_t2),
    by = "team2"
  ) %>%
    mutate(
    strength_diff = strength_t1 - strength_t2,
    avg_strength_diff = avg_strength_t1 - avg_strength_t2
  )

## Define feature set
feature_cols <- c(
  # ELO-based
  "elo_diff",
  
  # Player/squad strength
  "strength_diff",
  "avg_strength_diff",
  
  # Rolling form
  "win_rate_last5_t1", "win_rate_last5_t2",
  "win_rate_diff_last5",
  "win_rate_exp_t1", "win_rate_exp_t2",
  "win_rate_diff_exp",
  
  # Head-to-head
  "h2h_win_rate_t1", "h2h_advantage",
  
  # Venue
  "venue_win_rate_t1", "venue_win_rate_t2",
  
  # Scoring strength
  "avg_runs_scored_t1", "avg_runs_scored_t2",
  "avg_runs_conceded_t1", "avg_runs_conceded_t2",
  "net_scoring_diff",
  
  # Toss
  "toss_field_advantage",
  
  # Season context
  "season_win_rate_ytd_t1", "season_win_rate_ytd_t2"
)

model_input <- model_data %>%
  filter(season_year >= 2011) %>%
  select(all_of(c("match_id", "date", "season_year", feature_cols, "outcome"))) %>%
  drop_na(outcome, elo_diff)

# Walk-forward time split
## Train on seasons up to 2023, test on 2024 + 2025
split_year <- 2023

train_data <- model_input %>%
  filter(season_year <= split_year)

test_data <- model_input %>%
  filter(season_year > split_year & season_year < 2026)

# Recipe (preprocessing)
## Tidymodels recipe = declare transformations once, apply to train and test

ipl_recipe <- recipe(outcome ~ ., data = train_data) |>
  update_role(match_id, date, season_year, new_role = "ID") |>
  step_zv(all_predictors()) |>             # remove zero-variance FIRST
  step_nzv(all_predictors()) |>            # also remove near-zero-variance
  step_impute_median(all_numeric_predictors()) |>    # fill NAs with median
  step_normalize(all_numeric_predictors())           # z-score normalise last

# Models specifications
## Logistic regression
logit_spec <- logistic_reg(
  penalty = tune(), # L1/L2 via glmnet
  mixture = 1) %>%  # LASSO: forces uninformative features to zero
  set_engine("glmnet") %>%
  set_mode("classification")

## Random forest
rf_spec <- rand_forest(
  mtry = tune(),
  trees = 500,
  min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

## XGBoost
xgb_spec <- boost_tree(
  trees = 500,
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune(),
  sample_size = tune()) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

# Workflows
logit_wf <- workflow() %>%
  add_recipe(ipl_recipe) %>%
  add_model(logit_spec)

rf_wf <- workflow() %>%
  add_recipe(ipl_recipe) %>% 
  add_model(rf_spec)

xgb_wf <- workflow() %>% 
  add_recipe(ipl_recipe) %>% 
  add_model(xgb_spec)

# Cross-validation (5-fold on training data)
set.seed(12345)

cv_folds <- vfold_cv(train_data, v = 5, strata = outcome)
cv_metrics <- metric_set(roc_auc, mn_log_loss, accuracy, brier_class)

# Tune each model
tic("Logistic regression tuning")
logit_tune <- tune_grid(
  logit_wf,
  resamples = cv_folds,
  grid      = grid_regular(penalty(), levels = 20),
  metrics   = cv_metrics
)
toc()

tic("Random forest tuning")
rf_tune <- tune_grid(
  rf_wf,
  resamples = cv_folds,
  grid      = grid_latin_hypercube(
    mtry(range = c(3L, length(feature_cols))),
    min_n(),
    size = 15
  ),
  metrics = cv_metrics
)
toc()

tic("XGBoost tuning")
xgb_tune <- tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = grid_latin_hypercube(
    tree_depth(), learn_rate(), loss_reduction(), sample_prop(),
    size = 20
  ),
  metrics = cv_metrics
)
toc()

# Select best hyperparameters and fit final models
best_logit <- select_best(logit_tune, metric = "mn_log_loss")
best_rf    <- select_best(rf_tune,    metric = "mn_log_loss")
best_xgb   <- select_best(xgb_tune,  metric = "mn_log_loss")

final_logit <- finalize_workflow(logit_wf, best_logit) |> fit(train_data)
final_rf    <- finalize_workflow(rf_wf,    best_rf)    |> fit(train_data)
final_xgb   <- finalize_workflow(xgb_wf,  best_xgb)   |> fit(train_data)

# Evaluate on held-out test set
evaluate_model <- function(model, test, model_name) {
  preds <- augment(model, new_data = test)
  cv_metrics(
    preds,
    truth        = outcome,
    estimate     = .pred_class,
    .pred_team2_wins,          # first factor level — required by prob metrics
    event_level  = "first"
  ) |>
    mutate(model = model_name)
}

test_results <- bind_rows(
  evaluate_model(final_logit, test_data, "Logistic (GLM)"),
  evaluate_model(final_rf,    test_data, "Random Forest"),
  evaluate_model(final_xgb,   test_data, "XGBoost")
)

message("\n--- Test set performance ---")
test_results |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  arrange(mn_log_loss) |>
  print()

# Feature importance (from best model)
xgb_importance <- final_xgb |>
  extract_fit_parsnip() |>
  vip::vi() |>
  mutate(Variable = fct_reorder(Variable, Importance))

p_importance <- xgb_importance |>
  head(15) |>
  ggplot(aes(x = Variable, y = Importance)) +
  geom_col(fill = "#004BA0", width = 0.7) +
  coord_flip() +
  labs(
    title    = "XGBoost feature importance (top 15)",
    subtitle = "Gain-based importance",
    x        = NULL,
    y        = "Importance (gain)"
  ) +
  theme_minimal(base_size = 12)

############################### Final Prediction ################################
# Current ELO ratings using all played matches up to now
elo_final <- model_data %>%
  filter(has_result, !is.na(winner)) %>%
  arrange(date, match_id) %>%
  (\(df) {
    
    all_teams <- unique(c(df$team1, df$team2))
    elo_ratings <- setNames(rep(1500, length(all_teams)), all_teams)
    K <- 32
    
    walk(seq_len(nrow(df)), function(i) {
      row <- df[i, ]
      
      t1 <- row$team1
      t2 <- row$team2
      w  <- as.integer(row$winner == t1)
      
      ea <- 1 / (1 + 10 ^ ((elo_ratings[t2] - elo_ratings[t1]) / 400))
      
      elo_ratings[t1] <<- elo_ratings[t1] + K * (w - ea)
      elo_ratings[t2] <<- elo_ratings[t2] + K * ((1 - w) - (1 - ea))
    })
    
    tibble(
      team = names(elo_ratings),
      elo  = unname(elo_ratings)
    )
  })()

# Latest team form features
latest_features <- model_data %>%
  filter(!is.na(win_rate_exp_t1)) %>%
  arrange(date) %>%
  {
    bind_rows(
      select(., match_id, date, team = team1,
             win_rate_last5  = win_rate_last5_t1,
             win_rate_last10 = win_rate_last10_t1,
             win_rate_exp    = win_rate_exp_t1,
             venue_win_rate  = venue_win_rate_t1,
             avg_runs_scored = avg_runs_scored_t1,
             avg_runs_conceded = avg_runs_conceded_t1,
             season_win_rate_ytd = season_win_rate_ytd_t1),
      
      select(., match_id, date, team = team2,
             win_rate_last5  = win_rate_last5_t2,
             win_rate_last10 = win_rate_last10_t2,
             win_rate_exp    = win_rate_exp_t2,
             venue_win_rate  = venue_win_rate_t2,
             avg_runs_scored = avg_runs_scored_t2,
             avg_runs_conceded = avg_runs_conceded_t2,
             season_win_rate_ytd = season_win_rate_ytd_t2)
    )
  } %>%
  group_by(team) %>%
  slice_max(date, n = 1, with_ties = FALSE) %>%
  ungroup()

# Read remaining fixtures
remaining_fixtures <- read.csv("~/Desktop/IPL Predictive Model/Datasets/IPL_upcoming_fixture.csv") %>%
  mutate(
    match_id = as.character(match_id),
    season_year = as.integer(season_year),
    date = as.Date(date)
  )

# Prepare latest features for team1/team2
features_t1 <- latest_features %>%
  rename_with(~ paste0(.x, "_t1"), -team) %>%
  rename(team1 = team)

features_t2 <- latest_features %>%
  rename_with(~ paste0(.x, "_t2"), -team) %>%
  rename(team2 = team)

# Prepare squad strength for team1/team2
strength_pred_t1 <- squad_strength_2026 %>%
  rename(
    team1 = team,
    strength_t1 = squad_strength,
    avg_strength_t1 = avg_player_strength
  )

strength_pred_t2 <- squad_strength_2026 %>%
  rename(
    team2 = team,
    strength_t2 = squad_strength,
    avg_strength_t2 = avg_player_strength
  )

# Convert fixtures into model-ready rows
pred_rows <- remaining_fixtures %>%
  left_join(features_t1, by = "team1") %>%
  left_join(features_t2, by = "team2") %>%
  left_join(elo_final %>% rename(team1 = team, elo_team1 = elo), by = "team1") %>%
  left_join(elo_final %>% rename(team2 = team, elo_team2 = elo), by = "team2") %>%
  left_join(strength_pred_t1, by = "team1") %>%
  left_join(strength_pred_t2, by = "team2") %>%
  mutate(
    elo_diff = elo_team1 - elo_team2,
    strength_diff = strength_t1 - strength_t2,
    avg_strength_diff = avg_strength_t1 - avg_strength_t2,
    win_rate_diff_last5 = win_rate_last5_t1 - win_rate_last5_t2,
    win_rate_diff_exp = win_rate_exp_t1 - win_rate_exp_t2,
    h2h_win_rate_t1 = 0.5,
    h2h_advantage = 0,
    net_scoring_diff = (avg_runs_scored_t1 - avg_runs_conceded_t2) -
      (avg_runs_scored_t2 - avg_runs_conceded_t1),
    toss_field_advantage = 0,
    season_win_rate_ytd_t1 = season_win_rate_ytd_t1,
    season_win_rate_ytd_t2 = season_win_rate_ytd_t2,
    outcome = factor(NA, levels = levels(train_data$outcome))
  )

# Predict probabilities
pred_probs <- predict(final_xgb, new_data = pred_rows, type = "prob") %>%
  bind_cols(pred_rows %>% select(match_id, date, team1, team2, venue)) %>%
  select(match_id, date, team1, team2, venue, .pred_team1_wins, .pred_team2_wins) %>%
  rename(
    prob_team1 = .pred_team1_wins,
    prob_team2 = .pred_team2_wins
  )

# Monte Carlo winner probability simulation
N_SIMS <- 10000
set.seed(123)

# Current points table
current_points <- points_2026 %>%
  transmute(
    team = team,
    points = points,
    nrr = nrr,
    played = matches
  )

# Helper: simulate one match from model probability
simulate_match <- function(team1, team2, prob_team1) {
  if (runif(1) < prob_team1) team1 else team2
}

# Helper: get model probability for any playoff match
predict_playoff_prob <- function(team1_name, team2_name, venue_name = "Neutral") {
  
  playoff_fixture <- tibble(
    match_id = paste0("playoff_", sample(1e9, 1)),
    date = Sys.Date(),
    season_year = 2026L,
    team1 = team1_name,
    team2 = team2_name,
    venue = venue_name
  )
  
  playoff_row <- playoff_fixture %>%
    left_join(features_t1, by = "team1") %>%
    left_join(features_t2, by = "team2") %>%
    left_join(elo_final %>% rename(team1 = team, elo_team1 = elo), by = "team1") %>%
    left_join(elo_final %>% rename(team2 = team, elo_team2 = elo), by = "team2") %>%
    left_join(strength_pred_t1, by = "team1") %>%
    left_join(strength_pred_t2, by = "team2") %>%
    mutate(
      elo_diff = elo_team1 - elo_team2,
      strength_diff = strength_t1 - strength_t2,
      avg_strength_diff = avg_strength_t1 - avg_strength_t2,
      win_rate_diff_last5 = win_rate_last5_t1 - win_rate_last5_t2,
      win_rate_diff_exp = win_rate_exp_t1 - win_rate_exp_t2,
      h2h_win_rate_t1 = 0.5,
      h2h_advantage = 0,
      net_scoring_diff = (avg_runs_scored_t1 - avg_runs_conceded_t2) -
        (avg_runs_scored_t2 - avg_runs_conceded_t1),
      toss_field_advantage = 0,
      outcome = factor(NA, levels = levels(train_data$outcome))
    )
  
  predict(final_xgb, new_data = playoff_row, type = "prob") %>%
    pull(.pred_team1_wins)
}

# Prepare remaining match probabilities
remaining_match_probs <- pred_probs %>%
  select(match_id, date, team1, team2, prob_team1, prob_team2)

# Run simulations
sim_results <- map_dfr(seq_len(N_SIMS), function(sim) {
  
  # Start from current points
  pts <- current_points %>%
    select(team, points) %>%
    deframe()
  
  nrr_lookup <- current_points %>%
    select(team, nrr) %>%
    deframe()
  
  # Simulate remaining league matches
  for (i in seq_len(nrow(remaining_match_probs))) {
    
    game <- remaining_match_probs[i, ]
    
    winner <- simulate_match(
      team1 = game$team1,
      team2 = game$team2,
      prob_team1 = game$prob_team1
    )
    
    pts[winner] <- pts[winner] + 2
  }
  
  # Build final simulated league table
  final_table <- tibble(
    team = names(pts),
    points = as.numeric(pts),
    nrr = as.numeric(nrr_lookup[names(pts)])
  ) %>%
    arrange(desc(points), desc(nrr))
  
  top4 <- final_table %>%
    slice_head(n = 4) %>%
    pull(team)
  
  # -------------------------
  # Page Playoff System
  # -------------------------
  
  # Qualifier 1: 1st vs 2nd
  q1_prob <- predict_playoff_prob(top4[1], top4[2])
  q1_winner <- simulate_match(top4[1], top4[2], q1_prob)
  q1_loser <- setdiff(top4[1:2], q1_winner)
  
  # Eliminator: 3rd vs 4th
  elim_prob <- predict_playoff_prob(top4[3], top4[4])
  elim_winner <- simulate_match(top4[3], top4[4], elim_prob)
  
  # Qualifier 2: loser of Q1 vs winner of Eliminator
  q2_prob <- predict_playoff_prob(q1_loser, elim_winner)
  q2_winner <- simulate_match(q1_loser, elim_winner, q2_prob)
  
  # Final: winner of Q1 vs winner of Q2
  final_prob <- predict_playoff_prob(q1_winner, q2_winner)
  champion <- simulate_match(q1_winner, q2_winner, final_prob)
  
  tibble(
    sim = sim,
    champion = champion,
    top1 = top4[1],
    top2 = top4[2],
    top3 = top4[3],
    top4 = top4[4]
  )
})

# Title probabilities
title_probs <- sim_results %>%
  count(champion, name = "titles") %>%
  mutate(
    title_prob = titles / N_SIMS,
    team_colour = team_colours[champion]
  ) %>%
  arrange(desc(title_prob))

message("\n--- 2026 IPL Title Probabilities (Monte Carlo) ---")

title_probs %>%
  mutate(title_prob = percent(title_prob, accuracy = 0.1)) %>%
  select(champion, title_prob) %>%
  print()

# Top 4 probabilities
top4_probs <- sim_results %>%
  pivot_longer(cols = c(top1, top2, top3, top4),
               names_to = "position",
               values_to = "team") %>%
  count(team, name = "top4_finishes") %>%
  mutate(top4_prob = top4_finishes / N_SIMS) %>%
  arrange(desc(top4_prob))

message("\n--- 2026 IPL Top 4 Probabilities ---")

top4_probs %>%
  mutate(top4_prob = scales::percent(top4_prob, accuracy = 0.1)) %>%
  select(team, top4_prob) %>%
  print()
```

# 1. Executive Summary
This report delves into the predicting the winner of the 2026 Indian Premier League (IPL) season. The IPL is known to be the largest and most valuable cricket franchise league in the world, often recognised as the second-richest sports league globally in terms of per-match value, trailing only the NFL. The objective was to predict IPL match outcomes for the remaining fixture (post-match 24) using historical data of every IPL seasons, computing player ratings, developing machine learning models, and using monte-carlo simulation to quantify risk and uncertainty to produce the likelihood of which teams will be in top 4 and which team will win overall. The machine learning models that were developed were logistic regression, random forest, and XGBoost, out of which XGBoost illustrated strongest relative performance with an accuracy of 0.518, ROC AUC of 0.465, MN log loss of 0.694, and Brier score of 0.251. While predictive performance remains modest, reflecting the inherent volatility of T20 cricket, the model provides probabilistic insights rather than deterministic predictions.

Additionally, the most valuable predictors for the model in terms of gain-based importance, which is critical in demonstrating how much a feature improves prediction accuracy when used in splits, were:

- Current season performance (year-to-date win rates) 
- Scoring strength of each team (runs scored and conceded) 
- ELO rating (captures underlying team strength adjusted for opposition quality which is vital to differentiate a team's probability of beating strong teams vs weak teams)
- Recent team form (which showcases exponentially weighted recent win rates (recency-weighted win rates to shocase momentum). 

Monte Carlo simulation (10,000 iterations) was used to model the remainder of the season and playoff structure. The results suggest that the teams with the highest probability of winning the 2026 IPL season are:

- Punjab Kings: 28.2%
- Royal Challengers Bengaluru: 27.6%
- Sunrisers Hyderabad: 17.2%

In terms of playoff qualification, the most likely top 4 teams are:

- Royal Challengers Bengaluru: 94.1% 
- Punjab Kings: 93.4% 
- Sunrisers Hyderabad: 75.6% 
- Rajasthan Royals: 70.6%  

Overall, the analysis demonstrates that while individual match outcomes are highly uncertain, combining player-level modelling with team performance metrics and simulation techniques provides meaningful probabilistic insights into the tournament outcomes.

# 2. Data Overview
This analysis integrates multiple data sources to capture both match-level dynamics and underlying player and team strength. The primary dataset consists of historical IPL ball-by-ball data, which provides granular information on every delivery and enables the construction of match-level outcomes and performance metrics. This is supplemented with match-level summaries for efficient modelling, as well as 2026 season-specific data including squads, venues, and points tables. Additionally, auction price data is used as a proxy for player market valuation, thereby, enriching the modelling of player and team strength. 

Moreover, the match-level dataset serves as the core modelling unit as it provides information of every delivery ever bowled in IPL across all seasons. The player-level dataset is used for feature engineering, particularly in deriving batting and bowling performance ratings, which are aggregated to form team strength measures. Finally, the squad-level dataset links players to franchises, allowing dynamic estimation of team strength for the 2026 season, with the auction prices for each player being a proxy to balance player performance and quality (with higher selling price indicating higher demand and thereby, quality of player). 

To ensure data integrity, data quality checks were implemented. Missing values in key modelling variables were identified and handles through imputation or exclusion where appropriate. Season consistency was validated by correcting irregular season formats (e.g. split-year seasons like season 2020/2021), ensuring accurate temporal ordering for time-series feature engineering. Furthermore, team names were standardised across datasets to account for franchise naming inconsistencies, which is essential for accurate joins and aggregation. These steps ensure a robust and reliable dataset for exploratory data analysis and predictive modelling.

# 3. Exploratory Data Analysis (EDA)
EDA was conducted to identify the key factors influencing match outcomes and to inform the feature engineering process used in the predictive models. The analysis focused on uncovering structural patterns within the IPL, rather than purely descriptive trends.

The following segments in particular were explored:

- Historical team performance: to understand long-term franchise strength and dominance.
- Toss outcomes and match results: to evaluate whether winning the toss provides a systematic advantage.
- Scoring trends over time: to assess changes in batting conditions, especially in recent times where T20 is revolutionising cricket.
- Head-to-head dynamics: to identify persistent matchup advantages between teams.
- Venue effects: to determine whether teams benefit from playing at specific grounds (including exploring home ground advantage).

## 2.1 Historical Team Strength
Historical team strength is important to understand as this provides the context behind which franchise has won the most games over each season, and thereby, increasing the likelihood of them winning the championship. Teams with consistently strong performance across seasons are more likely to exhibit higher underlying quality in terms of squad composition, management, and tactical execution.

```{r, echo=FALSE}
p1
```

The graph above portrays how Mumbai Indians (MI) and Chennai Super Kings (CSK) have the top 2 highest all-time match wins since the beginning of IPL, corresponding to why they also have won the most championships (both teams have won 5 times each). This indicates that sustained success of particular teams are not random, but rather reflects the team's ongoing structural advantages. For example, one of the key reasons for this persistent success is both teams are known for retaining key valuable players over long periods. MI retain players like Rohit Sharma, Bumrah, Suryakumar Yadav, Hardik Pandya, and Tilak Varma, creating a stable core and maintaining team chemistry. CSK retained MS Dhoni, Shivam Dube, Ruturaj Gaikwad, Jadeja (excluding current season), Raina (earlier years), and Ashwin (earlier years). 

However, while historical performance provides useful context, it is not sufficient on its own for predicting future outcomes. The IPL is characterised by frequent player transfers, auction dynamics, and evolving team compositions, meaning that past success does not always translate directly into current strength. This partially explains why in the 2026 season, both MI and CSK are not doing as well as they have in previous years. From a modelling perspective, it's vital to incorporate dynamic features such as ELO ratings and recent form metrics to capture time-varing performance more effectively.

## Toss and Match Outcomes
The relationship between winning the toss and match outcomes provides insight into whether pre-match decisions lead to any advantages and thereby, resulting in a team winning the match. As shown in the figure below, the win rate of teams that win the toss has historically fluctuated between approximately 40% and 60% from 2008 to 2024, indicating that the toss alone has not been a consistently strong predictor of match outcomes. This suggests that match performance is driven more by team quality and in-game execution rather than the toss itself. 

```{r, echo = FALSE}
p2
```

However, a noticeable upward trend is observed in recent seasons, with the toss-winning team achieving win rates exceeding 60% in 2025 and the ongoing 2026 season. This shift may reflect evolving dyanmics in T20 cricket, including higher scoring environments and an increased preference for chasing, where teams benefit from knowing the target.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(gt)
library(scales)

# Build table
toss_bowl_pct <- model_data %>%
  filter(!is.na(toss_decision)) %>%
  group_by(season_year) %>%
  summarise(
    total_toss_wins = n(),
    bowl_first = sum(toss_decision == "field"),
    pct_bowl = bowl_first / total_toss_wins,
    .groups = "drop"
  ) %>%
  arrange(desc(season_year)) %>%
  filter(season_year >= 2020)

# Create formatted table
toss_bowl_pct %>%
  mutate(
    pct_bowl = percent(pct_bowl, accuracy = 0.1)
  ) %>%
  select(
    Season = season_year,
    `Total Matches` = total_toss_wins,
    `Chose to Bowl` = bowl_first,
    `% Bowled First` = pct_bowl
  ) %>%
  gt() %>%
  tab_header(
    title = "Toss Decisions by Season",
    subtitle = "Percentage of Teams Choosing to Bowl First After Winning the Toss"
  ) %>%
  cols_align(
    align = "center",
    everything()
  ) %>%
  tab_options(
    table.font.size = 12
  )
```

The table above further reinforces this where teams have preferred to bowl first, with proportion of team bowling first after winning the toss increasing from 64.3% in 2021 to 82.9% in 2025, and as high as 95.7% in the ongoing 2026 season. This indicates a clear and consistent strategic shift towards chasing, with teams increasingly preferring to field first. Improvements in pitch analysis, data availability, and tactical decision-making may also contribute to teams making more optimal choices after winning the toss.

It is important to note that the 2026 season is incomplete, and therefore these elevated win rates may be partially influenced by small sample effects. Nevertheless, the recent trend suggests that the toss may be becoming more strategically relevant, particularly in conditions that favour chasing. As a result, a toss-related feature is included in the model to capture this potential advantage, while recognising that its impact is likely to be secondary relative to team strength and form.

## Scoring Trends
The analysis of scoring trends provides insight into how the run-scoring environment in the IPL has evolved. The figure below illustrates a clear upward trend in average first innings totals over time, highlighting the evolving nature of T20 cricket. From 2008 to around 2018, average scores generally remained within the 140-170 range, whereas in recent seasons average totals have consistently increased, with many matches exceeding the 180-200 mark. This indicates a shift towards more batting-friendly conditions and increasingly aggressive scoring strategies. There is also a "pure business" aspect to this, with Muttiah Muralitharan (Sri Lankan legend cricketer currently appointed as SRH's spin bowling coach and mentor) stating that IPL "prepare flat pitches because low scores are boring for sponsors".

```{r, echo = FALSE}
p4
```

Furthermore, the increasing average runs scored in recent years, especially post-2022 is also due to the introduction of "impact player", which allows teams to substitute one member of their starting XI with a player from the substitutes list during the match. This is a tactical advantage during a team's batting and bowling innings as it gives them the chance to introduce a specialist batsman and specialist bowler and thereby, increase their depth in batting/bowling. This has played a substantial role in increasing the average scores per innings as well, which is showcased in the table below, where we see an increase in 200+ scores each season since, with the biggest jump being from the 2022 season and 2023 season (introduction of impact players).

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(gt)
library(scales)

# Build dataset (first innings only)
scores_200 <- deliveries_clean %>%
  filter(innings == 1) %>%
  group_by(match_id, season_year) %>%
  summarise(
    first_innings_total = max(team_runs, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(season_year >= 2022)

# Summarise by season
scores_summary <- scores_200 %>%
  group_by(season_year) %>%
  summarise(
    total_matches = n(),
    matches_200_plus = sum(first_innings_total >= 200),
    pct_200_plus = matches_200_plus / total_matches,
    .groups = "drop"
  ) %>%
  arrange(desc(season_year))

# Create formatted table
scores_summary %>%
  mutate(
    pct_200_plus = percent(pct_200_plus, accuracy = 0.1)
  ) %>%
  select(
    Season = season_year,
    `Total Matches` = total_matches,
    `200+ Scores` = matches_200_plus,
    `% 200+ Scores` = pct_200_plus
  ) %>%
  gt() %>%
  tab_header(
    title = "Frequency of 200+ First Innings Scores",
    subtitle = "IPL Seasons 2022–2026"
  ) %>%
  cols_align(
    align = "center",
    everything()
  ) %>%
  tab_options(
    table.font.size = 12
  )
```

From a modelling perspective, this reinforces the importance of incorporating scoring-based features, such as average runs scored, runs conceded, and net scoring differential. Teams that consistently generate higher totals, or restrict opponents effectively, have a significantly higher probability of winning. As such, scoring strength serves as a key predictor of match outcomes, capturing both offensive capability and defensive resilience.

## Head-to-Head Team Dynamics
The head-to-head win percentage matrix provides insight into historical matchup dynamics between teams, highlighting whether certain teams consistently outperform specific opponents. As shown in the figure below, while many matchups cluster around the 40-60% range, indicating a relatively balanced competition, there are several notable deviations where teams exhibit clear dominance. For example, GT have beaten SRH 5 out of the 6 games they have played, as seen by the 83% win rate. The high win rates suggest favourable matchups in batting depth, bowling composition, or tactical approaches. 

However, it is important to interpret these extremes with caution, as some high win percentages are based on relatively small sample sizes, which may overstate the strength of these relationships. Despite this, the presence of persistent imbalances across certain pairings indicates that head-to-head performance is not purely random and can reflect underlying strategic mismatches between teams. The findings justify the inclusions of head-to-head features in the models, such as historical win rates and relative matchup advantage, as they capture nuances not explained by overall team strength or recent form. While not the most dominant predictors, these features provide additional context that can improve model performance, particularly in closely matched games. 


## Venue/Home Advantage
Home advantage is often considered an important factor in cricket, as teams are expected to benefit from familiarity with pitch conditions, ground dimensions, and strong crowd support. However, the figure suggests that this advantage is not uniformly observed across IPL teams. The graph below delineates that since 2015, only Royal Challengers Bengaluru exhibit a home win rate above the 50% baseline (55%), while most other teams fall below this threshold, with several teams such as Mumbai Indians (17%) and Punjab Kings (30%) showing notably weak home performance.

```{r, echo=FALSE, fig.width=10, fig.height=8, out.width="100%"}
p5
```

This indicates that home advantage in the IPL is less pronounced than might be expected, and may be outweighed by other factors such as overall team strength, player form, and match conditions. One possible explanation is the standardisation of pitches and increased familiarity of players with different venues due to frequent travel and extensive match exposure. Additionally, the franchise-based structure and player mobility reduce the traditional notion of “home-ground specialists,” as players often have experience playing across multiple venues.

From a modelling perspective, these findings suggest that while venue-specific performance may still provide useful contextual information, it is unlikely to be a dominant predictor of match outcomes. Instead, venue effects are incorporated through features such as venue win rates, but are treated as secondary to stronger predictors like team form, scoring strength, and ELO ratings.

# 4. Feature Engineering
Feature engineering transforms raw match and player data into structured variables that capture the underlying drivers of match outcomes. In the context of T20 cricket, this involves quantifying team strength, recent performances, scoring ability, and contextual factors like venues. The objective is to construct features that reflect both short-term momentum and long-term quality, enabling the model to distinguish between strong and weak teams in a given matchup.

## Team Form Features
Team form features were constructed using each team's historical match results in chronological order. Each match was converted into a binary result for each team:

$$
\text{won} = \begin{cases}
  1, & \text{if the team won} \\
  0, & \text{if the team lost}
\end{cases}
$$
Rolling win rates were then calculated over recent matches (last 5 and 10 matches) to showcase short-term momentum. Importantly, the current match was excluded using a lag, preventing look-ahead bias. 

An exponentially weighted win rate was also constructed to give more weight to recent results. This was calculated using decay weights, where the most recent match receives the highest weight and older matches receive progressively lower weights. A decay factor of $\lambda = 0.85$ was used, meaning recent matches influence the feature more strongly than older matches.

In simple terms: 

$$
\textbf{Exp Weighted Win Rate} = \frac{\sum w_i \times \text{win}_i}{\sum w_i}
$$

where $w_i = \lambda^i$, and older matches receive smaller weights.

This feature captures short-term momentum more effectively than a simple average. For example, a team that has won 3 recent matches after a poor start will receive a stronger form signal than a team with the same overall record but poorer recent results.

## Head-to-Head Features
Head-to-head features were created to capture opponent-specific matchup effects. For each team-opponent pair, the model calculated the team's historical win rate against that opponent prior to the current match. This was constructed cumulatively:

$$
\textbf{H2H Win Rate} = \frac{\text{Previous wins against opponent}}{\text{Previous matches against opponent}}
$$
To avoid leakage, only matches played before the current match were used. Where no prior head-to-head matches existed, the win rate was set to 0.5, representing a neutral prior. This feature is useful because some teams may consistently match up well against others due to tactical matchups or venues. However, it was interpreted cautiously because some matchups have small sample sizes, which can exaggerate apparent dominance.

## Venue Features
Venue features were designed to capture whether teams perform better at particular ground. For each team-venue combination, the model calculated the team's historical win rate at that venue prior to the current match. 

The feature was calculated as: 
$$
\textbf{Venue Win Rate} = \frac{\text{Previous wins at venue}}{\text{Previous matches at venue}}
$$
A minimum threshold of 3 previous matches was applied before assigning a venue win rate for credibility and thereby, reduced noise from very small samples. If a team had fewer than 3 matches at a venue, he value was treated as missing and later handled through median imputation in the modelling recipe. This feature reflects factors such as pitch conditions, ground dimensions, crowd support, and familiarity with local playing conditions.

## ELO Ratings
ELO ratings were used to estimate dynamic team strength over time. Each team started with an initial rating of 1500, and ratings were updated after every completed match. Before each match, the expected win probability for Team 1 was calculated using the difference in ratings:

$$
E_A = \frac{1}{1 + 10^{(R_B - R_A)/400}}
$$

where $R_A$ and $R_B$ are the ELO ratings of the two teams.

After the match, the winning team gained rating points and the losing team lost rating points:

$$
R_A^{\text{new}} = R_A + K(S_A - E_A)
$$
where:

- $S_A = 1$ if Team A won, and 0 otherwise
- $E_A$ is Team A's expected win probability
- $K = 32$ controls how quickly ratings update

ELO is useful because it adjusts for opponent strength. Beating a strong team increases a team's rating more than beating a weaker team, while losing to a weak team is penalised more heavily. The main model input was `elo_diff`, calculated as `elo_diff` = `elo_team1` - `elo_team2`, with a positive value indicating Team 1 had a stronger pre-match rating.

## Scoring Strength
Scoring strength features were constructed from innings-level run totals. The aim was to capture both batting power and bowling effectiveness. For each team, rolling averages were calculated for:

- Average runs scored in recent innings
- Average runs conceded in recent innings

These were based on previous matches only, again avoiding look-ahead bias. The key derived metric was net scoring differential:
$$
\text{Net Scoring Diff} = \text{(Team 1 Runs Scored} - \text{Team 2 Runs Conceded)} - \text{(Team 2 Runs Scored} - \text{Team 1 Runs Conceded)}
$$
This feature compares how Team 1's attacking strength matches up against Team 2's defensive record, and vice cersa. It essentially looks at whether Team 1 is expected to outscore Team 2 based on recent scoring and conceding patterns. This is especially important in modern IPL cricket (and T20 cricket), where high scores have become more frequent and run-scoring ability is a major determinant of match outcomes.

## Player-Level Modelling
Player ratings were developed to account for changes in squad composition between seasons. This is important because IPL teams change significantly after auctions each year, meaning historical team performance alone may not accurately reflect current team strength.

Batting ratings were constructed using:

- total runs
- batting average
- strike rate

A raw batting score was calculated using standardised versions of these metrics. Standardisation was used so that metrics on different scales could be combined fairly. A credibility adjustment was then applied based on balls faced to overcome inflated batting scores:
$$
\textbf{Batting Credibility} = \min\left(\frac{\text{balls faced}}{300}, 1\right)
$$
This prevents small-sample players from having an inflated rating. For example, a tailender who hits one six from one ball would have a very high strike rate, but low credibility, so their batting rating would be heavily reduced. 

Bowling ratings were constructed using:

- economy rate
- bowling average
- bowling strike rate
- wickets taken

Since lower economy, lower average, and lower strike rate indicate better bowling, these metrics were multiplied by negative weights after standardisation. A credibility adjustment was also applied based on overs bowled:
$$
\textbf{Bowling Credibility} = \min\left(\frac{\text{overs bowled}}{50}, 1\right)
$$
This again prevents part-time bowlers from receiving extreme ratings based on very small bowling samples. The final performance rating was: $\textbf{Performance Rating} = \textbf{Batting Rating} + \textbf{Bowling Rating}$.

## Auction Price Adjustment
Auction price was incorporated as a market-based proxy for player quality. The intuition is that auction prices reflect not only historical performance, but also expected future value, scarcity of role, and franchise demand. Since auction prices can be highly skewed, a log transformation was applied with $\textbf{Auction Rating} = \log(1 + \text{Sold Price})$. This was then standardised and blended with the player performance rating: 
$$
\text{Final Player Rating} = 0.7 \times \text{Performance Rating} + 0.3 \times \text{Auction Rating}
$$
This weighting ensures that realised performance remains the primary driver, while auction price acts as a supplementary prior. This is useful for players with limited historical IPL data but high market valuation.

## Squad Strength
Squad strength was calculated by linking the final player ratings to the 2026 squad dataset. For each franchise, players were ranked by final rating, and the top players were aggregated to estimate current team strength. The main squad strength metric was:
$$
\textbf{Squad Strength} = \sum \text{Final Player Ratings}
$$

A squad strength difference was then calculated for each match:
$$
\textbf{Strength Diff} = \text{Squad Strength}_{Team1} - \text{Squad Strength}_{Team2}
$$
This feature directly addresses the issue of IPL squad turnover. Instead of assuming that a franchise's past performance fully represent its current strength, the model incorproates the estimated quality of its current players. 

Overall, the feature engineering process combined historical match outcomes, recent form, scoring ability, venue context, opponent-specific dynamics, and player-level squad strength. This allowed the model to capture both realised team performance and underlying player quality. The final feature set therefore reflects the key drivers of IPL outcomes: form, team strength, scoring dominance, matchup effects, and current squad composition.

# 5. Modelling
The objective of the modelling framework is to estimate the probability of Team 1 winning a given IPL match. This is formulated as a binary classification problem, where the outcome variable takes the value 1 if Team 1 wins and 0 if Team 2 wins. The dataset was constructed at the match level, with each observation represented a completed match and containing engineered features capturing team form, scoring strength, ELO ratings, venue effects, head-to-head performance, and squad strength.

To ensure robust model evaluation, the data was split chronologically into training and testing sets, with the training set being every season before 2023 (inclusive), and test set being all seasons 2024 onwards. This preserves the time-series nature of the data and avoids look-ahead bias, ensuring that models are evaluation on their ability to generalise to future matches rather than memorising past outcomes.

## Models Implemented
### Logistic Regression
Logistic regression serves as a baseline model, estimating the probability of Team 1 winning as a function of a linear combination of features:
$$P(\text{Team 1 wins} \mid \mathbf{x}) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p)}}$$
L1 regularisation (LASSO) was applied via the `glmnet` engine, with the penalty parameter $lamda$ tuned over a grid of 20 values. LASSO is particularly well-suited to this setting as it performs implicit feature selection by shrinking uninformative information coefficients to exactly zero, yielding an interpretable model, with the surviving coefficients represented as log-odds relativities. 

### Random Forest
A random forest ensemble of 500 decision trees was fitted using the `ranger` engine. Random forests extend single decision trees by constructing many trees in parallel, each fitted on a bootstrap sample of the training data with a random subset of predictors considered at each split. The final prediction is the majority vote (or average probability) across all trees, reducing variance relative to any individual tree. This approach is robust to multicollinearity, a relevant property given the natural correlation between rolling form features and ELO ratings, and captures non-linear interactions between predictors without requiring explicit specification. The number of candidate predictors at each split (`mtry`) and the minimum node size (`min_n`) were jointly tuned using a Latin hypercube search across 15 configurations, with `mtry` finalised against the post-recipe predictor count to prevent the model from attempting splits on more variables than the existing ones in the processed data.

### XGBoost
Gradient boosting via XGBoost was the third model implemented, using the `xgboost` engine within the `tidymodels` framework. Unlike random forests, which build trees independetly in parallel, XGBoost constructs trees sequentially, with each tree fitted to the residuals of the current ensemble. This iterative error-correction mechanism typically yields superior predictive performance on tabular data with heterogenous feature types. The regularisation framework within XGBoost also penalises tree complexity, reducing the risk of overfitting on a dataset of moderate size. Four hyperparameters were jointly optimised via a 20-point Latin hypercube search: tree depth, learning rate, minimum loss reduction required for a node split, and the row subsampling proportion per tree.

## Preprocessing Pipeline
A unified `tidymodels` recipe was applied consistently to both training and test partitions, ensuring that no information from the test set influenced any preprocessing decisions. The steps were applied in the following order:

1. Zero and near-zero variance removal: features with negligible variance across the training set were dropped prior to any scaling. This step was applied first because feeding a zero-variance predictor (such as the binary indicator variable of teams having an advantage from winning the toss and choosing to field first which exhibited no variation in certain cross-validation folds) into a normalisation step produces undefined scaling behaviour and can corrupt downstream model fitting.

2. Median imputation: missing values arising from insufficient match history (e.g. rolling win rates for early-career seasons, venue rates below the 3 game credibility threshold) are imputed with the training set median.

3. Z-score normalisation: all remaining numeric predictors are standardised to zero mean and unit variance. While tree-based models are scale-invariant, normalisation was retained uniformly to ensure logistic regression receives comparably scaled inputs and to simplify coefficient interpretation.

## Model Evaluation 
All three models were evaluated on the held-out 2024-2025 test set across four metrics: accuracy, log-loss (primary selection criterion), ROC-AUC, and Brier score. Log-loss was prioritised as the selection criterion because it is a proper scoring rule that penalises both miscalibrated probability and incorrect classification, which is particularly important in a simulation context where the model's probability outputs, rather than point predictions, are consumed by the Monte Carlo engine.

The table shows that all three models perform similarly, with no single model dominant across all metrics. XGBoost achieves the best ROC-AUC (0.504), indicating the strongest ability to discriminate between match outcomes, while Logistic regression records the best log-loss (0.694) and Brier score (0.250), reflecting marginally superior probability calibration. Random Forest trails on all probability metrics with log-loss of 0.717 and Brier score of 0.261, suggesting its probability estimates are less reliable despite matching Logistic regression on accuracy (51.8%).

Despite logistic regression's marginal edge on log-loss, XGBoost was selected as the model for final predictions. This was because XGBoost's superior ROC-AUC means its probability ranking are more reliable, which is critical when win probabilities feed directly into the Monte Carlo simulation where small differences in estimated probabilities compound across thousands of iterations. Additionally, XGBoost's capacity to capture non-linear interaction between features such as ELO ratings, rolling form, and scoring strength provides a structural advantage over logistic regression, which is constrained to linear decision boundaries. The marginal log-loss difference of 0.001 between the two models is negligible and does not offset this advantage.

```{r echo=FALSE, message=FALSE, warning=FALSE}
model_results <- test_results |>
  select(model, .metric, .estimate) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  rename(
    Model         = model,
    Accuracy      = accuracy,
    `ROC-AUC`     = roc_auc,
    `Log-Loss`    = mn_log_loss,
    `Brier Score` = brier_class
  ) |>
  arrange(`Log-Loss`)

model_results |>
  gt() |>
  tab_header(
    title    = "Model Performance on Held-Out Test Set (2024–2025)",
  ) |>
  fmt_number(
    columns = c(Accuracy, `ROC-AUC`, `Log-Loss`, `Brier Score`),
    decimals = 3
  ) |>
  cols_align(
    align   = "center",
    columns = c(Accuracy, `ROC-AUC`, `Log-Loss`, `Brier Score`)
  ) |>
  data_color(
    columns = c(`Log-Loss`, `Brier Score`),
    colors  = scales::col_numeric(
      palette = c("#d4edda", "#ffffff", "#f8d7da"),
      domain  = NULL
    )
  ) |>
  data_color(
    columns = c(Accuracy, `ROC-AUC`),
    colors  = scales::col_numeric(
      palette = c("#f8d7da", "#ffffff", "#d4edda"),
      domain  = NULL
    )
  ) |>
  tab_source_note(
    source_note = "↓ lower is better for Log-Loss and Brier Score; ↑ higher is better for Accuracy and ROC-AUC"
  )
```

The ROC-AUC values for all three models are approximately 0.46–0.49, indicating that the models exhibit limited discriminative ability, performing close to a random classifier (AUC = 0.5), as seen from the graph below. This suggests that, based on the available pre-match features, the models struggle to consistently distinguish between winning and losing teams.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(pROC)

roc_data <- bind_rows(
  augment(final_logit, test_data) |> mutate(model = "Logistic (GLM)"),
  augment(final_rf,    test_data) |> mutate(model = "Random Forest"),
  augment(final_xgb,   test_data) |> mutate(model = "XGBoost")
)

roc_curves <- roc_data |>
  group_by(model) |>
  summarise(
    roc = list(roc(
      response  = outcome,
      predictor = .pred_team2_wins,              # team2_wins as positive class
      levels    = c("team1_wins", "team2_wins"), # negative first, positive second
      direction = "<"
    )),
    .groups = "drop"
  ) |>
  mutate(auc = map_dbl(roc, ~round(as.numeric(auc(.x)), 3)))

roc_plot_data <- roc_curves |>
  mutate(
    tpr = map(roc, ~.x$sensitivities),
    fpr = map(roc, ~1 - .x$specificities)
  ) |>
  select(model, tpr, fpr, auc) |>
  unnest(c(tpr, fpr))

# Build label positions dynamically — sorted by AUC, evenly spaced
label_data <- roc_curves |>
  select(model, auc) |>
  arrange(desc(auc)) |>
  mutate(
    label_y     = seq(0.22, 0.08, length.out = n()),  # evenly spaced regardless of n models
    label_x     = 0.62,
    label_text  = glue::glue("{model} (AUC = {auc})")
  )

p_roc <- roc_plot_data |>
  ggplot(aes(x = fpr, y = tpr, colour = model)) +
  geom_line(linewidth = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_text(
    data          = label_data,
    aes(x = label_x, y = label_y, label = label_text, colour = model),
    size          = 3.5,
    fontface      = "bold",
    show.legend   = FALSE,
    inherit.aes   = FALSE
  ) +
  scale_colour_manual(values = c(
    "Logistic (GLM)" = "skyblue",
    "Random Forest"  = "plum",
    "XGBoost"        = "springgreen4"
  )) +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0.01)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0.01)) +
  labs(
    title    = "**ROC Curves of all 3 Models**",
    subtitle = "Dashed diagonal = random classifier baseline",
    x        = "False positive rate (1 - Specificity)",
    y        = "True positive rate (Sensitivity)",
    colour   = NULL
  ) +
  theme_ipl()

p_roc
```

This result is expected in the context of T20 cricket, which is inherently high-variance. Match outcomes can be significantly influenced by unpredictable in-game events such as dropped catches, no-balls, or short bursts of individual brilliance, none of which can be captured using pre-match information. The narrow spread of performance across logistic regression, random forest, and XGBoost further indicates that model architecture is not the primary constraint on predictive performance. Instead, the limiting factor lies in the feature set, suggesting that additional predictive power would likely require richer features, such as real-time match conditions, player availability, or more granular contextual data. Despite the modest ROC-AUC values, the models remain useful for generating probabilistic forecasts, which can be leveraged in simulation frameworks to estimate tournament-level outcomes.

## Feature Importance
Feature importance provides essential information in which predictors contribute more and less to the machine learning model's prediction. The XGBoost feature importance plot highlights that current season performance and scoring-related metrics are the dominant drivers of match outcomes. In particular, season-to-date win rate and average runs scored exhibit the highest gain-based importance, indicating that teams performing well in the ongoing season and consistently generating high scores have a significantly higher probability of winning. This reinforces the importance of recent form and offensive strength in the T20 format, where momentum and run-scoring ability are critical. 

Scoring balance, captured through net scoring differential and runs conceded, also ranks highly. This suggests that match outcomes are not solely driven by batting strength, but by a team's ability to both score efficiently and restrict opponents. The present of ELO rating among the top features further indicates that underlying team quality, adjusted for opponent strength, remains an important baseline predictor, complementing short-term performance indicators. Mid-tier features such as exponentially weighted win rates and head-to-head performance provide additional context, capturing recent momentum and matchup-specific dynamics. However, their relatively lower importance suggests that while these factors contribute to prediction, they are secondary to broader indicators of team strength and scoring capability.

Notably, venue-related features and squad strength differences appear among the least important predictors. This aligns with earlier exploratory findings that home advantage in the IPL is inconsistent, and suggests that pre-match structural strength (as captured by squad ratings) may already be reflected in performance-based features such as win rates and scoring metrics. Overall, the model places greatest emphasis on what teams are doing now (form and scoring) rather than where they are playing or historical structural assumptions about team strength.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p_importance
```

# 6. Monte Carlo Simulation
Having established the XGBoost model as the preferred predictive engine, the remaining 2026 IPL season was simulated using a Monte Carlo framework across 10,000 iterations. This approach treats each remaining fixture as a probabilistic event, with match-win probabilities generated by the XGBoost model, and spreads uncertainty across the full league stage and playoff bracket to produce distributional estimates of each team's championship probability. 

## League Stage Simulation
For each simulation, the current points table served as the starting state, with all remaining league fixtures resolved stochastically using the XGBoost-predicted match probabilities. Two points were awarded for each simulated win, with the final simulated points table (sorted by points and net run rate as a tiebreaker) determining the top four teams that qualify for the playoffs.

## Playoff Structure
The IPL employs the Page playoff system, which provides the top 2 finishers with a double chance. In each simulation, the 4 playoff teams were resolved as follows:

- Qualifier 1: 1st vs 2nd where the winner advances directly to the final
- Eliminator: 3rd vs 4th where the loser is eliminated
- Qualifier 2: Loser of Qualifier 1 vs winner of Eliminator where the winner advances to the final
- Final: Winner of Qualifier 1 vs winner of Qualifier 2

Playoff match probabilities were generated dynamically using the XGBoost model, incorporating each team's most recent form features and current ELO ratings at the time of the playoff, rather than assuming fixed probabilities throughout. This ensures that the simulation updates appropriately as the league stage resolves.

## Simulation Results
```{r echo=FALSE, message=FALSE, warning=FALSE}
title_probs %>%
  left_join(current_points, by = c("champion" = "team")) %>%
  arrange(desc(title_prob)) %>%
  select(
    Team = champion,
    `Title Probability` = title_prob,
  ) %>%
  mutate(`Title Probability` = scales::percent(`Title Probability`, accuracy = 0.1)) %>%
  gt() %>%
  tab_header(
    title = "2026 IPL Championship Probabilities",
    subtitle = "Monte Carlo simulation (10,000 iterations) · XGBoost match win probabilities"
  ) %>%
  cols_align(align = "center", columns = -Team)
```

Punjab Kings and Royal Challengers Bengaluru emerge as the joint title favourites at 28.2% and 27.6% respectively, reflecting their strong current-season form and scoring metrics which are the features the model weights most heavily. Sunrisers Hyderabad (17.2%) represent a credible third contender, supported by their consistently aggressive batting approach which scores highly on the scoring strength features. The remaining teams' title probabilities are modest, reflecting either current points deficits that are difficult to overcome with remaining fixtures, or weaker form ratings entering the simulation.

```{r echo=FALSE, message=FALSE, warning=FALSE}
top4_probs %>%
  select(
    Team = team,
    `Top 4 Probability` = top4_prob
  ) %>%
  mutate(`Top 4 Probability` = scales::percent(`Top 4 Probability`, accuracy = 0.1)) %>%
  gt() %>%
  tab_header(
    title = "2026 IPL Playoff Qualification Probabilities",
    subtitle = "Probability of finishing in the top 4 across 10,000 simulations"
  ) %>%
  cols_align(align = "center", columns = -Team) %>%
  tab_options(table.font.size = 12)
```

In terms of playoff qualification, RCB (94.1%) and Punjab Kings (93.4%) are near-certainties to reach the top 4, while Sunrisers Hyderabad (75.6%) and Rajasthan Royals (70.6%) occupy the more contested third and fourth positions.

# 7. Limitations and Future Enhancements
While the model does a great job at providing probabilistic insight into IPL outcomes, there are several limitations, including:

1. The inherent stochasticity of T20 cricket means that a substantial portion of match outcomes lies beyond the reach of any pre-match model, with the irreducible variance in a 20 over format being structurally higher than in longer formats.

2. Squad strength is currently treated as a static feature, which does not reflect within-season player form, injury status, or other player rotation policies. To overcome this limitation, making this time-vering would be a meaningful enhancement (i.e. by weighting player ratings against their 2026 ball-by-ball performance).

3. The playoff probability calibration is directly contingent on the accuracy of remaining fixture win probabilities. As the season progresses and more matches are played, the model's rolling features will become increasingly representative of the current team, and re-running the simulation after each match day is recommended.

Potential future enhancements include probability calibration using monotonic regression (as input X increases, prediction Y never decreases) to address the overconfidence observed in the random forest, ensemble stacking of the 3 models to extract complementary signal, and the incorporation of weather and pitch report.

# 8. Conclusion
This report developed a data-driven framework to predict IPL match outcomes and estimate tournament-winning probabilities by integrating historical performance, team form, scoring strength, and player-level features. While the models demonstrated limited predictive accuracy—reflecting the inherently high uncertainty of T20 cricket—they provided meaningful probabilistic insights into team performance and relative strength. The results highlight that recent form and scoring ability are the most critical determinants of success, while factors such as venue advantage and historical matchups play a secondary role. By embedding these insights within a Monte Carlo simulation framework, the analysis extends beyond match-level predictions to quantify each team’s likelihood of reaching the playoffs and ultimately winning the IPL. Overall, this approach illustrates that even in a highly volatile sporting environment, structured modelling can provide valuable strategic insights, with probabilities offering a more realistic and informative perspective than deterministic predictions.
